---
title: "Sarah: Spatial Statistics Project Models"
output: html_document
date: "2024-11-24"
---

# Preprocessing

### Load Libraries
```{r, message=FALSE, warning=FALSE}
library(sp)
library(sf)
library(rstan)
library(rstanarm)
library(spdep)
library(dplyr)
library(spatialreg)
library(ggplot2)
```

### Import datasets
```{r, message=FALSE, warning=FALSE}
unique_stores <- read.csv("https://raw.githubusercontent.com/sarahvastani/newrepo/refs/heads/main/unique_stores.csv")
Traffic_estimate_Socio_econ <- read.csv("https://raw.githubusercontent.com/sarahvastani/newrepo/refs/heads/main/Traffic_estimate_Socio_econ.csv")
cleaned_Store_Demographic <- read.csv("https://raw.githubusercontent.com/sarahvastani/newrepo/refs/heads/main/cleaned_Store_Demographic_new.csv")
expected_customer_traffic_withProxmity <- read.csv("https://raw.githubusercontent.com/sarahvastani/newrepo/refs/heads/main/expected_customer_traffic_withProxmity.csv")
```

### Merge datasets
```{r, message=FALSE, warning=FALSE}
merged_data <- merge(cleaned_Store_Demographic, unique_stores, by = "Store_name", all = FALSE)
merged_data_2 <- merge(expected_customer_traffic_withProxmity, Traffic_estimate_Socio_econ, by = "StoreID", all = FALSE)
merged_data_2 <- merged_data_2 %>%
  rename(Store_name = StoreID)
merged_data_combined <- merge(merged_data, merged_data_2, by = "Store_name")
```

### Aggregate Data
```{r, message=FALSE, warning=FALSE}
aggregated_data <- merged_data_combined %>%
  group_by(Store_name, Latitude, Longitude) %>%
  summarise(
    Utility_Socio = mean(Utility_Socio, na.rm = TRUE),
    Utility_Traffic = mean(Utility_Traffic, na.rm = TRUE),
    Distance = mean(Distance, na.rm = TRUE),
    Estimated_Traffic = mean(Estimated_Traffic, na.rm = TRUE),
    Mean_Utility = mean(Mean_Utility, na.rm = TRUE),
    .groups = "drop"
  )
```

### Convert to sf object
```{r, message=FALSE, warning=FALSE}
spatial_data <- st_as_sf(aggregated_data, coords = c("Longitude", "Latitude"), crs = 4326)

```

### Create spatial neighbors based on the distance between points
```{r, message=FALSE, warning=FALSE}
coords <- st_coordinates(spatial_data)  
neighbors <- dnearneigh(coords, d1 = 0, d2 = 1)  
```

### Create a spatial weights matrix from the neighbors
```{r, message=FALSE, warning=FALSE}
weights <- nb2listw(neighbors, style = "W")
```

### Standardizing predictors
```{r, message=FALSE, warning=FALSE}
spatial_data$Latitude <- st_coordinates(spatial_data)[,2]
spatial_data$Longitude <- st_coordinates(spatial_data)[,1]
spatial_data$Estimated_Traffic <- scale(spatial_data$Estimated_Traffic)
spatial_data$Latitude <- scale(spatial_data$Latitude)
spatial_data$Longitude <- scale(spatial_data$Longitude)
spatial_data$Utility_Socio <- scale(spatial_data$Utility_Socio)
spatial_data$Utility_Traffic <- scale(spatial_data$Utility_Traffic)
spatial_data$Distance <- scale(spatial_data$Distance)
spatial_data$Mean_Utility <- scale(spatial_data$Mean_Utility)
```

# MODEL 1

### Run a Bayesian spatial model using rstanarm
```{r, echo=FALSE, message=FALSE, warning=FALSE}

model_bayesian <- stan_glm(
  Estimated_Traffic ~ Latitude + Longitude + Utility_Socio + Utility_Traffic + Distance + Mean_Utility,
  data = spatial_data,
  family = gaussian(),
  prior = normal(0, 5), 
  chains = 4,
  cores = 4, 
  iter = 2000,
  warmup = 1000
)

```

### Check model summary
```{r, message=FALSE, warning=FALSE}
summary(model_bayesian)
```

### Posterior samples plot
```{r, message=FALSE, warning=FALSE}
plot(model_bayesian)
```

### Compute Mean Squared Error (MSE) 
```{r, message=FALSE, warning=FALSE}
predicted_values <- predict(model_bayesian, type = "response")
actual_values <- spatial_data$Estimated_Traffic
mse <- mean((predicted_values - actual_values)^2)
cat("MSE: ", mse, "\n")
```

### Compute R-squared
```{r, message=FALSE, warning=FALSE}
sst <- sum((actual_values - mean(actual_values))^2)
sse <- sum((actual_values - predicted_values)^2)
r_squared <- 1 - (sse / sst)
cat("R²: ", r_squared, "\n")
```
### Plot Actual vs. Predicted Traffic
```{r, message=FALSE, warning=FALSE}
comparison_table_bayesian <- data.frame(
  Predicted_Traffic = predicted_values,
  Actual_Traffic = actual_values
)

plot_bayesian <- ggplot(comparison_table_bayesian, aes(x = Predicted_Traffic, y = Actual_Traffic)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE, linetype = "dashed") +
  labs(
    title = "Bayesian Model: Actual vs. Predicted Traffic",
    subtitle = paste("MSE:", round(mse, 3), "| R²:", round(r_squared, 3)),
    x = "Predicted Traffic",
    y = "Actual Traffic"
  ) +
  theme_minimal()

print(plot_bayesian)
```

# MODEL 2

### Fit a spatial lag model using the spatial weights matrix with adjusted tol.solve and method
```{r, message=FALSE, warning=FALSE}
model_gaussian <- lagsarlm(
  Estimated_Traffic ~ Latitude + Longitude + Utility_Socio + Utility_Traffic + Distance + Mean_Utility,
  data = spatial_data,
  listw = weights,
  tol.solve = 1e-12,  
  method = "Matrix"   
)
```

### Check model summary
```{r, message=FALSE, warning=FALSE}
summary(model_bayesian)
```

### Posterior samples plot
```{r, message=FALSE, warning=FALSE}
plot(model_bayesian)
```

### Compute Mean Squared Error (MSE) 
```{r, message=FALSE, warning=FALSE}
predicted_values_gaussian <- predict(model_gaussian, newdata = spatial_data, listw = weights)
actual_values <- spatial_data$Estimated_Traffic
mse_gaussian <- mean((predicted_values_gaussian - actual_values)^2)
cat("MSE: ", mse_gaussian, "\n")
```

### Compute R-squared
```{r, message=FALSE, warning=FALSE}
sst_gaussian <- sum((actual_values - mean(actual_values))^2)
sse_gaussian <- sum((actual_values - predicted_values_gaussian)^2)
r_squared_gaussian <- 1 - (sse_gaussian / sst_gaussian)
cat("R²: ", r_squared_gaussian, "\n")
```

### Plot Actual vs. Predicted Traffic
```{r, message=FALSE, warning=FALSE}
comparison_table_gaussian <- data.frame(
  Predicted_Traffic = as.numeric(predicted_values_gaussian),
  Actual_Traffic = as.numeric(actual_values)
)

plot_gaussian <- ggplot(comparison_table_gaussian, aes(x = Predicted_Traffic, y = Actual_Traffic)) +
  geom_point(color = "green", alpha = 0.6) +
  geom_smooth(method = "lm", color = "orange", se = FALSE, linetype = "dashed") +
  labs(
    title = "Gaussian Model: Actual vs. Predicted Traffic",
    subtitle = paste("MSE:", round(mse_gaussian, 3), "| R²:", round(r_squared_gaussian, 3)),
    x = "Predicted Traffic",
    y = "Actual Traffic"
  ) +
  theme_minimal()

print(plot_gaussian)
```


